---
title: "In-class Exercise Week 4"
author: "Cher In Teo"
format: html
date: "September 18, 2026"
date-modified: today
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
  message: false
---

## Load Packages

```{r}
pacman::p_load(sf, ggstatplot, spdep, tmap, tidyverse, knitr, GWmodel)
```

## Data Import & Management

Import Hunan shapefile & CSV file

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

Join both and tidy the data table and filter columns using representative numbers instead of typing the column names to reduce typos

```{r}
hunan_sf <- left_join(hunan,hunan2012) %>%
  select(1:3, 7, 15, 16, 31, 32)
```

## Map GDPPC

Trial and error to adjust the dimensions.

Code hidden:

```{r}
#| fig-width: 16
#| fig-height: 6
#| echo: false
basemap <- tm_shape(hunan_sf) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan_sf, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

Code shown:

:::{style="font-size: 1.50em}
```{r}
#| fig-width: 16
#| fig-height: 6
#| eval: false
basemap <- tm_shape(hunan_sf) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan_sf, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

## Convert to SpatialPolygonDataFrame

```{r}
hunan_sp <- hunan_sf %>%
  as_Spatial()
```

## Geographically Weighted Summary Statistic (GSWW) with adaptive bandwidth

Determine adaptive bandwidth

```{r}
bw_CV <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "CV",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

```{r}
bw_CV
```

```{r}
bw_AIC <- bw.gwr(GDPPC ~1,
                data = hunan_sp,
                approach = "AICc",
                adaptive = TRUE,
                kernel = "bisquare",
                longlat = T)
```

```{r}
bw_AIC
```

Computing Geographically Weighted Summary Statistics

The code chunk below computes gwss (GWmodel package). 

The analysis is performed using adaptive distance weighted.

The distance is derived using AICc method and the kernal method used is bisquare.

```{r}
gwstat <- gwss(data = hunan_sp,
               vars = "GDPPC",
               bw = bw_AIC,
               kernel = "bisquare",
               quantile = TRUE,
               adaptive = TRUE,
               longlat = T)
```

Note: gwss supports 5 kernel methods - gaussian, exponential, bisquare, tricute and boxcar. Try the rest.

## Prepare Output Data'

This only works if no sorting has been applied prior. This method cannot be used if data has been sorted before.

```{r}
gwstat_df <- as.data.frame(gwstat$SDF)
```

Next, cbind() is used to append the newly derived data.frame onto hunan_sf sf data.frame.

```{r}
hunan_gstat <- cbind(hunan_sf, gwstat_df)
```

## Visualising Geographically Weighted Summary Statistics

```{r}
tm_shape(hunan_gstat) +
  tm_polygons(fill = "GDPPC_LM",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
                fill.legend = tm_legend(
                  title = "local Mean")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title ("Distribution of Geographically Weighted Mean") +
  tm_layout(legend.position = c("left","bottom"))
```


              )
  

