---
title: "Geospatial Data Science with R"
format: html
---

Geospatial Data Science is the process of importing, wrangling, integrating, and processing geographically referenced data sets.

## Load Package

```{r}
pacman::p_load(sf, tidyverse)
```

## Import Polygon Feature Data (geojson)

```{r}
mpsz = st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.geojson")
```

## Import Polygon Feature Data (shapefile)

```{r}
cyclingpath = st_read(dsn = "data/geospatial/CyclingPath_Mar2026", 
                       layer = "CyclingPathGazette")
```

## Data Checks

### st_geometry()

Report shows basic information of the feature class (e.g. geometry type, geographic extent of features, coordinate system of data)

```{r}
st_geometry(mpsz)
```

### glimpse()

Report shows data type of each field.

```{r}
glimpse(mpsz)
```

### head()

Report shows complete information of a feature object and allows number of records to be selected.

```{r}
head(mpsz, n=5)
```

## Plotting the Geospatial Data (Visualisation)

Multiplot of all attributes:

```{r}
plot(mpsz)
```

With warning suppressed:

```{r}
#| warning: false
plot(mpsz)
```

Geometry-only plot:

```{r}
plot(st_geometry(mpsz))
```

Plot with specific attribute:

```{r}
plot(mpsz["PLN_AREA_N"])
```

Plot preschool layer on top of mpsz layer:

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

## Projection Transformation

To perform geoprocessing using two geospatial datasets, both datasets must be projected using the same coordinate reference system.

### Update ESPG Code

This is used to correct missing or wrongly assigned coordinate system during the data import process.

To display currently assigned coordinate system:

```{r}
st_crs(mpsz)
```

The above indicates that EPSG is 4326 (degree of latitude or longtitude).

To change EPSG to 3414 (measure distance):

```{r}
mpsz <- st_transform(mpsz, 3414)
```

To verify EPSG has changed:

```{r}
st_crs(mpsz)
```

The same is done for preschool which was downloaded in geojson format as well.

Check current coordinate system:

```{r}
st_geometry(preschool)
```

```{r}
preschool <- st_transform(preschool, 3414)
```

To verify EPSG has changed:

```{r}
st_crs(preschool)
```

Note: EPSG 3414 is the EPSG code number for the SVY21 coordinate reference system.

## Import Aspatial Data

Examples of aspatial data are .csv files.

Create a tibble data frame:

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

### list()

Create a list with the tibble data frame:

```{r}
list(listings)
```

## Convert to Spatial Object

```{r}
listings_sf <- st_as_sf(listings, 
               coords = c("longitude", "latitude"),
               crs=4326) %>%
st_transform(crs = 3414)
```

Note: coords - x = longitude and y = latitude in standard mapping convention.\
crs - requires coordinates system in EPSG format. 4326 is wgs84 Geographic Coordinate System and EPSG: 3414 is Singapore SVY21 Projected Coordinate System. %\>% - nest st_transform() to transform the newly created simple feature data frame into svy21 projected coordinates system. (feed to next function)

Review transformation:

```{r}
glimpse(listings_sf)
```

Plot listings_sf layer on top of mpsz layer:

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(listings_sf), add = TRUE)
```

## Geoprocessing with sf package

### Use Case 1: Land Acquisition Analysis

Scenario: The authority is planning to upgrade the exiting cycling path. To do so, they need to acquire 5 metres of reserved land on the both sides of the existing cycling path. You are tasked to determine the extend of the land need to be acquired and their total area.

Solution:

1.  Compute the 5-meter buffers around cycling paths.

```{r}
buffer_cycling <- st_buffer(
  cyclingpath, dist=5, nQuadSegs = 30)
```

```{r}
buffer_cycling <- st_transform(buffer_cycling, 3414)
```

2.  Calculate the area of the buffers:

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

3.  Derive total land area impacted.

```{r}
sum(buffer_cycling$AREA)
```

4.  Create a plot to show the buffer by using a selected planning subzone (i.e. Tampine West).

<!-- -->

i)  Extract polygon feature of Tampines West:

```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST") 
```

ii) Clip cycling buffers within Tampines West planning subzone.

```{r}
buffer_cycling_selected <- st_intersection(
  buffer_cycling, mpsz_selected)
```

iii) Create the plot.

```{r}
#| warning: false
plot(st_geometry(buffer_cycling_selected))
```

### Use Case 2: Determine Number of Preschools by Planning Subzone

Scenario: The authority requires a count of pre-schools for each Planning Subzone to support forward planning. Using R and the sf package, perform the necessary geoprocessing to compute these counts and present the results clearly.

Solution:

1.  Identify and calculate number of preschools in each Planning Subzone:

```{r}
mpsz$`PreSchool Count`<- lengths(st_intersects(mpsz, preschool))
```

Check summary statistics:

```{r}
summary(mpsz$`PreSchool Count`)
```

2.  List Planning Subzone with most preschools:

```{r}
top_n(mpsz, 1, `PreSchool Count`)
```

3.  Derive area of each Planning Subzone:

```{r}
mpsz$Area <- mpsz %>%
  st_area()
```

4.  Compute Preschool Density:

```{r}
mpsz <- mpsz %>%
  mutate(`PreSchool Density` = `PreSchool Count`/Area * 1000000)
```

5.  Data Visualisation

<!-- -->

i)  Plot histogram to reveal distribution of Preschool Density. Basic histogram:\

```{r}
hist(mpsz$`PreSchool Density`)
```

```         
Improved visuals:
```

```{r}
ggplot(data=mpsz, 
       aes(x= as.numeric(`PreSchool Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are Preschools Evenly Distributed in Singapore?",
       subtitle= "There are many planning subzones with only 1 preschool and 7 planning subzones with at least 30 or more preschools",
      x = "Preschool Density (per km sq)",
      y = "Frequency")
```

ii) Scatterplot to show relationship between Preschool Density and Preschool Count:

```{r}
#| warning: false
ggplot(data=mpsz, 
       aes(y = `PreSchool Count`, 
           x= as.numeric(`PreSchool Density`)))+
  geom_point(color="black", 
             fill="light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Preschool Density (per km sq)",
      y = "Preschool Count")
```
